$p$-adic precision

### 2014-02-06 padicprec.tex
Precision on a multiple root of a polynomial.

Suppose that we are given a monic polynomial $P \in K[X]$ written as a 
product $P_0 = A_0 B_0$ where $A_0$ and $B_0$ are monic and coprime. 
Hensel's lemma implies that there exists a small neighborhood $\mathcal 
U$ of $P_0$ such that any monic polynomial $P \in \mathcal U$ factors 
uniquely as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ 
and $B_0$ respectively. Thus, we can consider the application $f : P 
\mapsto (A,B)$ defined on the subset of $\mathcal U$ consisting of monic 
polynomials. We want to differentiate $f$ at $P_0$. For this, we 
differentiate the equality $P = A B$ around $P_0$, obtaining
\begin{equation}
\label{eq:difffactor}
dP = A_0 \cdot dB + B_0 \cdot dA.
\end{equation}
where $dP$, $dA$ and $dB$ have degree less than $\deg P$, $\deg A$ and 
$\deg B$ respectively. If $A_0 U_0 + B_0 V_0 = 1$ is a Bezout relation
between $A_0$ and $B_0$, it follows from Eq.~\eqref{eq:difffactor} that
$dA$ (resp. $dB$) is the remainder in the Euclidean division of $V_0
{\cdot} dP$ by $A_0$ (resp. of $U_0 {\cdot} dP$ by $B_0$).

An important special case of the previous study occurs when $A_0$ has 
degree $1$, that is $A_0(X) = X - \alpha$ with some $\alpha \in K$. The 
application $P \mapsto A$ is then nothing but the mapping that follows 
the simple root $\alpha$. Of course, its differential around $P_0$ can 
be computed by the above method. Nevertheless there is a more direct way 
to obtain it in this particular case: we write $(P + dP)(\alpha + 
d\alpha) = 0$ and expand this. We obtain this way $P'(\alpha) d\alpha + 
dP(\alpha) = 0$ where $P'$ denotes the derivative of $P$. Since $\alpha$ 
is a simple root, $P'(\alpha)$ does not vanish and we recover $d \alpha 
= - \frac{dP(\alpha)}{P'(\alpha)}$.

\todo{Discuss a bit multiple roots.}

>>>>

Suppose that we are given a polynomial $P_0 \in K_d[X]$ written as a 
product $P_0 = A_0 B_0$ where $A_0$ and $B_0$ are monic and coprime. 
Hensel's lemma implies that there exists a small neighborhood $\mathcal 
U$ of $P_0$ in $K_d[X]$ such that any $P \in \mathcal U$ factors 
uniquely as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ 
and $B_0$ respectively. Thus, we can consider the application $f : P 
\mapsto (A,B)$ defined on the subset of $\mathcal U$ consisting of monic 
polynomials. We want to differentiate $f$ at $P_0$. For this, we 
differentiate the equality $P = A B$ around $P_0$, obtaining
\begin{equation}
\label{eq:difffactor}
dP = A_0 \cdot dB + B_0 \cdot dA.
\end{equation}
where $dP$, $dA$ and $dB$ have degree less than $\deg P$, $\deg A$ and 
$\deg B$ respectively. If $A_0 U_0 + B_0 V_0 = 1$ is a Bezout relation
between $A_0$ and $B_0$, it follows from Eq.~\eqref{eq:difffactor} that
$dA$ (resp. $dB$) is the remainder in the Euclidean division of $V_0
{\cdot} dP$ by $A_0$ (resp. of $U_0 {\cdot} dP$ by $B_0$).

\subsubsection*{Finding roots}

An important special case of the previous study occurs when $A_0$ has 
degree $1$, that is $A_0(X) = X - \alpha_0$ with some $\alpha_0 \in K$. 
The application $P \mapsto A$ is then nothing but the mapping that 
follows the simple root $\alpha_0$. Of course, its differential around 
$P_0$ can be computed by the above method. Nevertheless there is a more 
direct way to obtain it in this particular case: we write $(P_0 + 
dP)(\alpha_0 + d\alpha) = 0$ and expand this. We obtain this way 
$P_0'(\alpha_0) d\alpha + dP(\alpha_0) = 0$ where $P'$ denotes the 
derivative of $P$. Since $\alpha_0$ is a simple root, $P_0'(\alpha_0)$ 
does not vanish and we find:
$$d \alpha = - \frac{dP(\alpha_0)}{P'_0(\alpha_0)}.$$

We now address the case of a multiple roots. Let $P_0$ be a monic 
polynomial of degree $d$ and $\alpha_0 \in K$ be a root of $P_0$ having 
multiplicity $m > 1$. Due to the multiplicity, it is no longer possible 
to follow to root $\alpha_0$ in a neighborhood of $P_0$. But it is if we 
restrict ourselves to polynomials which have a root of multiplicity $m$ 
close to $\alpha_0$. More precisely, let us consider the locus $\mathcal 
V_m$ consisting of monic polynomials of degree $d$ having a root of 
multiplicity at least $m$. It is a Zariski close subset of $K_d[X]$ and 
it contains by assumption the polynomial $P_0$ we started with. 
Moreover, the irreducible components of $\mathcal V_m$ that meet at 
$P_0$ are in bijection with the set of roots of $P_0$ of multiplicity 
$\geq m$. In particular $\alpha_0$ correspond to one of these 
irreducible components; let us denote it by $\mathcal V_{m,\alpha_0}$. 
The algebraic variety $\mathcal V_{m,\alpha_0}$ is a \emph{a fortioti} 
$K$-manifold. Moreover there exists a differentiable function $f$ which 
is defined on a neighborhood of $P_0$ in $\mathcal V_{m,\alpha_0}$ and 
follows the root $\alpha_0$, \emph{i.e.} $f(P_0) = \alpha_0$ and for all 
$P$ such that $f(P)$ is defined, $\alpha = f(P)$ is a root of 
multiplicity (at least) $m$ of $P$. The existence of $f$ follows from 
Hensel's Lemma applied to the factorisation $P_0(X) = (X-\alpha_0)^m 
B_0(X)$ where $B_0(X)$ is a polynomial which is coprime to $X - 
alpha_0$. We can finally compute the differential of $f$ at $P_0$ 
(along $\mathcal V_{m,\alpha_0}$) by differentiating the equality
$P^{(m-1})(\alpha) = 0$ where $P^{(i})$ denotes the $i$-th derivative
of $P$. We find:
\begin{equation}
\label{eq:dalphamult}
d \alpha = - \frac{dP^{(m-1)}(\alpha_0)}{P_0^{(m)}(\alpha_0)}.
\end{equation}

The above computation has actually interesting consequences on $p$-adic 
precision. Let us illustrate this with an exemple. Consider the monic 
polynomial $P(X) = X^2 + O(p^{2N}) X + O(p^{2N})$ (where $N$ is some 
large integer) and assume that we are asking a softwate to compute one 
of its roots. What is the right answer? We remark that, if $\alpha$ is 
any $p$-adic number divisible by $p^n$, then $P(X)$ can be $X^2 - 
\alpha^2$ whose roots are $\pm \alpha$. Conversely, we can prove that 
the two roots of $P$ are necessarily divisible by $p^n$. Hence, the
right answer is certainly $O(p^N)$. Nevertheless, if we know in addition 
that $P$ has a double root, say $\alpha$, we can write $P(X) = (X - 
\alpha)^2$ and identifying the coefficient in $X$, we get $\alpha =
O(p^{2N})$ (or $O(p^{2N-1})$ if $p=2$). 
This is exactly the result we get by applying Lemma \ref{lem:main} and
using Eq.~\eqref{eq:dalphamult} to simplify the differential.
